---
title: "Getting started with datasetviewer"
vignette: >
  %\VignetteIndexEntry{Getting started with datasetviewer}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format: html
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`datasetviewer` renders a dataset in a fast, scrollable grid modelled on the
SAS Studio table viewer: a column-selection panel, a per-column property
pane (Label, Name, Length, Type, Format, Informat), a names-versus-labels
header toggle, header-driven sort, and a free-text row filter. The same
widget runs in interactive Shiny apps and in static HTML documents.

The data is sent to the browser as Parquet and queried in place with
DuckDB-WASM, so filtering, sorting, and scrolling stay fast on large
datasets without sampling rows.

## A plain data frame

```{r}
library(datasetviewer)
dataset_viewer(mtcars)
```

Right-click a column header to sort or add a filter. Click the funnel in the
toolbar to edit the filter expression directly, for example `mpg > 20 and
cyl == 4`.

## Column metadata and labels

When the `artoo` package is installed, column metadata (labels, formats,
informats, storage lengths) is read from a labelled or artoo-conformed
frame and shown in the property pane. Switch the **View** dropdown to
*Column labels* to use labels as headers.

```{r}
#| eval: false
library(artoo)
adsl <- read_dataset("adsl.parquet")
dataset_viewer(adsl, view = "labels")
```

## In Shiny

`datasetviewerOutput()` and `renderDatasetViewer()` embed the widget, and the
current filter, sort, and column selection are published back as inputs
(`input$<id>_filter`, `input$<id>_sort`, `input$<id>_columns`,
`input$<id>_view`) so the app can reuse them server-side.

```{r}
#| eval: false
library(shiny)
ui <- fluidPage(datasetviewerOutput("viewer", height = "500px"))
server <- function(input, output, session) {
  output$viewer <- renderDatasetViewer(dataset_viewer(mtcars))
}
shinyApp(ui, server)
```
