---
title: "Get started with datasetviewer"
vignette: >
  %\VignetteIndexEntry{Get started with datasetviewer}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format: html
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Looking at a dataset should not mean choosing between *fast* and *complete*.
Grids that render every cell into the page bog down at a few thousand rows;
ones that page through a server add a network round-trip to every scroll and
filter. The usual escape hatch -- show the first 1,000 rows -- quietly hides
exactly the rows you opened the viewer to find.

`datasetviewer` takes a different route, borrowed from the
[SAS Studio table viewer](https://documentation.sas.com/doc/en/webeditorcdc/v_001/webeditorug/p1pr6m3i1p7n4nn19f8e8x3yw1xq.htm).
The data is sent to the browser **once**, as Parquet, and queried in place
with [DuckDB-WASM](https://duckdb.org/docs/api/wasm/overview); the grid is
drawn on an HTML canvas and only ever materialises the rows you can see. Sort,
filter, hide a column, jump to the last page -- each is a SQL query over the
whole dataset that returns in milliseconds, with **no row sampling**. The same
widget runs in an interactive Shiny app and in a static HTML document like
this one.

## Your first viewer

Hand `dataset_viewer()` a data frame. That is the entire API for the common
case -- everything else is interaction inside the widget.

```{r}
library(datasetviewer)
dataset_viewer(mtcars)
```

Try it: drag the scrollbar, drag a column border to resize, or click a row.
The grid above is live -- it is the real widget, not a screenshot.

## A guided tour of the interface

The layout mirrors SAS Studio, so anyone who has used that viewer is already
at home:

- **Columns panel** (left) -- a checklist of every column with a type chip
  (`A` for character, `#` for numeric, a calendar for dates). Uncheck a column
  to hide it from the grid; the data is never reloaded.
- **Property pane** (lower left) -- select a column to inspect its `Label`,
  `Name`, `Length`, `Type`, `Format`, and `Informat`, the same attributes
  `PROC CONTENTS` reports.
- **View toggle** (toolbar) -- switch the header row between variable
  **names** and **labels**.
- **Header menu** -- right-click any column header to sort ascending or
  descending, add a filter, copy the column, or size the columns to content.
- **Status bar** -- the total row and column counts, and the filtered count
  once a filter is active.

::: {.callout-tip}
Sorting and filtering are driven from the widget, not from R arguments, so a
reader of your report can explore the data themselves without re-running any
code.
:::

## CDISC metadata, labels, and the property pane

A plain data frame has no labels, so the property pane shows names only. Point
the viewer at a labelled or CDISC-conformed frame and the metadata comes to
life. With the companion [`artoo`](https://vthanik.github.io/artoo/) package
installed, column labels, formats, informats, and storage lengths are read
straight from the frame and shown in the property pane -- and you can set the
header row to use labels instead of names.

```{r}
#| eval: !expr requireNamespace("artoo", quietly = TRUE)
# artoo ships the CDISC pilot ADaM datasets used across these docs.
dataset_viewer(artoo::cdisc_adsl, view = "labels")
```

Select `STUDYID` in the columns panel and the property pane reads
*Study Identifier*; the header row now shows labels because of
`view = "labels"`. No `artoo` dependency is required for plain frames -- it is
consulted only when present.

::: {.callout-note}
`dataset_viewer()` also accepts a **path** to a dataset file
(`dataset_viewer("adsl.parquet")`); `artoo::read_dataset()` reads it, carrying
its metadata into the property pane.
:::

## Filtering the whole table

There are two ways to filter, both operating over every row:

1. **Filter Table Rows** -- click the funnel in the toolbar and type a
   free-text expression, SAS-style, such as `AGE >= 75 and SEX = "F"`. It is
   translated to a SQL `WHERE` clause, and the status bar updates to the
   matched count.
2. **Add Filter** -- right-click a column header. The dialog adapts to the
   column's type: a checklist of distinct values for character columns, a
   comparison operator and value for numbers, and a date picker for dates.

Because the filter runs in DuckDB over the full Parquet payload, the answer is
exact -- the matched count is the true count, not a count within a sampled
window.

## Built for scale: no row sampling

The design choice that makes this work is moving the query engine into the
browser:

- **Transport.** The frame is serialised to Parquet with `nanoparquet` and
  carried in the widget payload -- columnar, compressed, and read natively by
  the engine.
- **Engine.** DuckDB-WASM reads that Parquet directly and answers every
  filter, sort, and page as SQL.
- **Grid.** The canvas grid asks the engine only for the rows in the visible
  window (a `LIMIT`/`OFFSET` query), so scrolling cost is independent of the
  dataset's size.

The practical upshot: a viewer over a multi-million-row frame scrolls, sorts,
and filters as smoothly as one over `mtcars`, and every row stays reachable.

## Embedding in a Shiny app

In Shiny, pair `datasetviewerOutput()` in the UI with `renderDatasetViewer()`
on the server. The viewer is not a dead end: the user's current column
selection, filter, sort, and view mode flow **back** into the app as inputs,
namespaced by the output id, so the rest of the app can react to what the
analyst is looking at.

```{r}
#| eval: false
library(shiny)
library(datasetviewer)

ui <- fluidPage(
  datasetviewerOutput("viewer", height = "560px"),
  verbatimTextOutput("state")
)

server <- function(input, output, session) {
  output$viewer <- renderDatasetViewer(dataset_viewer(mtcars))

  # State changes in the widget arrive as inputs, namespaced by output id.
  output$state <- renderPrint({
    list(
      columns = input$viewer_columns, # columns currently shown
      filter  = input$viewer_filter, # active filter expression
      sort    = input$viewer_sort, # active sort
      view    = input$viewer_view # "names" or "labels"
    )
  })
}

shinyApp(ui, server)
```

## Static HTML and Quarto

No server is required for the static case -- this very vignette embeds live
widgets. Drop `dataset_viewer()` into any R Markdown or Quarto document and
the result is a self-contained, fully interactive grid that works offline once
loaded. The same call you would write in a Shiny app produces the same viewer
here.

## Where to next

- `?dataset_viewer` -- the full argument reference, including `view`, `width`,
  and `height`.
- `?datasetviewerOutput` and `?renderDatasetViewer` -- the Shiny bindings and
  the input names the widget publishes.
- [`artoo`](https://vthanik.github.io/artoo/) -- lossless CDISC dataset I/O
  and the metadata model the property pane reads.
